Shapes analysis for time series.

### 2024-10-14 NEURIPS24/paper.tex
update

%Reproducing Kernel Hilbert Space (RKHS)
\vspace{-1ex}
\section{Background on LDDMM}
\vspace{-1ex}
\label{section:LDDMM}
\sam{Shape-FPCA employs the Square Root Velocity Field (SRVF) representation to separate space and time. However, this method is designed for continuous objects and only applies to time series of the same length. This issue can be addressed through interpolation, but this approach is not always reliable in sparse and irregular sampling scenarios. Most FDA papers [2,3,4] that we are aware of cope with the same issue using interpolation or basis function expansion. In a nutshell, FDA methods have to deal with continuous objects, while LDDMM algorithms can keep a discrete-to-discrete analysis. }
In this part, we expose how to
 learn the diffeomorphisms $(\phi_j)_{j\in[N]}$ using LDDMM, initially introduced in \cite{beg2005computing}.
 In a nutshell, for any $j\in [N]$, $\phi_j$ corresponds to a differential flow related to a learnable velocity field belonging to a well-chosen Reproducing Kernel Hilbert Space (RKHS).

@@

\sam{Lien avec functional data analysis ? Insister sur le fait que concaténer temps et espace c'est moisi}

@@

 \vspace{-1ex}
    \paragraph{Choice for the kernel associated with the RKHS $\msv$}
    \label{paragraph:kernel_V}
    As depicted on \Cref{fig:diffeo}-\ref{fig:transport}, we can not use any kernel $K$ to apply the previous methodology to learn deformations on time series' graphs.
    We describe and motivate our choice in this paragraph.
     Denote the one-dimensional Gaussian kernel by $K_\sigma^{(a)}(x,y)=\exp(-|x-y|^2/\sigma)$ for any $(x,y)\in (\Rset^a)^2$, $a\in \Nset$ and $\sigma>0$.
    To solve the geodesic shooting problem \eqref{eq:relaxation} on $\Rset^{d+1}$, we consider for $\msv$ the RKHS associated with the kernel defined for any $(t,x),(t',x')\in (\Rset^{d+1})^2$:
    \begin{align}
      \label{eq:kernel_TAS}
      &K_{\msg}((t,x),(t',x'))=\begin{pmatrix}
        c_0K_{\text{time}} & 0 \\
        0 & c_1 K_{\text{space}} 
        \end{pmatrix} \eqsp ,\\
       & K_{\text{space}}=K_{\sigma_{T,1}}^{(1)}(t,t')K_{\sigma_x}^{(d)}(x,x') \Idd\eqsp,K_{\text{time}}=K_{\sigma_{T,0}}^{(1)}(t,t') \eqsp,
    \end{align}
    parametrized by the widths $\sigma_{T,0},\sigma_{T,1},\sigma_x>0$ and the constants $c_0,c_1>0$.
This choice for $K_\msg$ is motivated by the representation \Cref{theorem:representation} and the following result. 
    \begin{lemma}
      \label{lemma:choice_of_kernel_V}
      If we denote by $\msv$ the RKHS associated with the kernel $K_{\msg}$, then for any vector field $v$ generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0},
       there exist $\gamma \in \msd(\Rset) $ and $f\in \rmC^1(\Rset^{d+1},\Rset^d)$ such that $\phi^v=\Psi_\gamma\circ\Phi_f $.
    \end{lemma}
    Instead of Gaussian kernels, other types of smooth kernels can be selected as long as the structure \eqref{eq:kernel_TAS} is respected.
    % \begin{proof}
    %   Let $v$ be a vector field generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0}.
    %  We remark that the first coordinate of the velocity field $v_\tau$ denoted by $v_\tau^{\text{time}}$ only depends on the time variable $t$ for any $\tau\in[0,1]$.
    %  Thus, when computing the first coordinate of the deformation $\phi^v$, denoted by $\gamma$, we integrate \eqref{eq:LDDMM_dynamic} with $v_\tau$ replaced by $v_\tau^{\text{time}}$,
    %   thus $\gamma$ is independant of the variable $x$. Moreover, $\gamma\in \mcd(\Rset)$ since a Gaussian kernel induced an Hilbert space $\msv$ satisfying $|f|_V\leq |f|_\infty+ |\dd f|_\infty  $ for any $f\in \msv$ by \citep[Theorem 9]{glaunes2005transport}.
    %   For the same reason, we have $\phi^v\in \mcd(\Rset^{d+1})$, and thus its last coordinates denoted by $f$ belongs to $\rmC^1(\Rset^{d+1},\Rset^d)$, and by construction $\phi^v=\Psi_\gamma\circ\Phi_f $.
    % \end{proof
    \begin{remark}
      With this choice of kernel, the features associated with the time transformation can be extracted from the momentums $(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0}$ in \eqref{eq:def_v0} by taking the coordinates related to time.
      However, the features related to the space transformation are not only in the space coordinates since the related kernel $K_{\text{space}}$ depends on time as well.
      \end{remark}
      In \Cref{appendix:kernel_TS_LDDMM}, we give guidelines for selecting the hyperparameters $(\sigma_{T,0},\sigma_{T,1},\sigma_x,c_0,c_1)$.

@@

 LDDMM framework has weaker assumptions than the Metamorphosis framework \cite{blanz2003face}. The 3DMM framework requires that each mesh be re-parametrized into a consistent form where the number of vertices, triangulation, and the anatomical meaning of each vertex are consistent across all meshes, as stated in the introduction of \cite{booth20163d}.
 In our context, we do not need such pre-processing; the time series graph can have different sizes.

 Leaving the LDDMM representation, the results of \cite{srivastava2010shape,heo2024logistic} address the representation of curves with the Square-Root Velocity (SRV) representation.
 However, the SRV representation is applied after reparametrization of the temporal dimension of the unit length segment.
 Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
 Very recently, in a functional data analysis framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
 Nevertheless, this method is made for \textit{continuous objects} and only applies to time series of \textit{same length}, making the estimation more sensitive to noise.

@@

 All these state-of-the-art methods previously mentionned \cite{glaunes2008large,oh2024stable,wu2024shape,heo2024logistic} are compared to TS-LDDMM in \Cref{appendix: robustness} and \Cref{appendix: shape_classification}.
 \vspace{-1ex}
  \section{Limitations and conclusion}
  \vspace{-1ex}
  \label{sec:limitations}

>>>>

%Reproducing Kernel Hilbert Space (RKHS)
\vspace{-1ex}
\section{Background on LDDMM}
\vspace{-1ex}
\label{section:LDDMM}

In this part, we expose how to
 learn the diffeomorphisms $(\phi_j)_{j\in[N]}$ using LDDMM, initially introduced in \cite{beg2005computing}.
 In a nutshell, for any $j\in [N]$, $\phi_j$ corresponds to a differential flow related to a learnable velocity field belonging to a well-chosen Reproducing Kernel Hilbert Space (RKHS).

@@

%\sam{Lien avec functional data analysis ? Insister sur le fait que concaténer temps et espace c'est moisi}

@@

    \paragraph{Choice for the kernel associated with the RKHS $\msv$}
    \label{paragraph:kernel_V}
    As depicted on \Cref{fig:diffeo}-\ref{fig:transport}, we can not use any kernel $K$ to apply the previous methodology to learn deformations on time series' graphs.
    We describe and motivate our choice in this paragraph.
     Denote the one-dimensional Gaussian kernel by $K_\sigma^{(a)}(x,y)=\exp(-|x-y|^2/\sigma)$ for any $(x,y)\in (\Rset^a)^2$, $a\in \Nset$ and $\sigma>0$.
    To solve the geodesic shooting problem \eqref{eq:relaxation} on $\Rset^{d+1}$, we consider for $\msv$ the RKHS associated with the kernel defined for any $(t,x),(t',x')\in (\Rset^{d+1})^2$:
    \begin{align}
      \label{eq:kernel_TAS}
      &K_{\msg}((t,x),(t',x'))=\begin{pmatrix}
        c_0K_{\text{time}} & 0 \\
        0 & c_1 K_{\text{space}} 
        \end{pmatrix} \eqsp ,\\
       & K_{\text{space}}=K_{\sigma_{T,1}}^{(1)}(t,t')K_{\sigma_x}^{(d)}(x,x') \Idd\eqsp,K_{\text{time}}=K_{\sigma_{T,0}}^{(1)}(t,t') \eqsp,
    \end{align}
    parametrized by the widths $\sigma_{T,0},\sigma_{T,1},\sigma_x>0$ and the constants $c_0,c_1>0$.
This choice for $K_\msg$ is motivated by the representation \Cref{theorem:representation} and the following result. 
    \begin{lemma}
      \label{lemma:choice_of_kernel_V}
      If we denote by $\msv$ the RKHS associated with the kernel $K_{\msg}$, then for any vector field $v$ generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0},
       there exist $\gamma \in \msd(\Rset) $ and $f\in \rmC^1(\Rset^{d+1},\Rset^d)$ such that $\phi^v=\Psi_\gamma\circ\Phi_f $.
    \end{lemma}
    Instead of Gaussian kernels, other types of smooth kernels can be selected as long as the structure \eqref{eq:kernel_TAS} is respected.
    % \begin{proof}
    %   Let $v$ be a vector field generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0}.
    %  We remark that the first coordinate of the velocity field $v_\tau$ denoted by $v_\tau^{\text{time}}$ only depends on the time variable $t$ for any $\tau\in[0,1]$.
    %  Thus, when computing the first coordinate of the deformation $\phi^v$, denoted by $\gamma$, we integrate \eqref{eq:LDDMM_dynamic} with $v_\tau$ replaced by $v_\tau^{\text{time}}$,
    %   thus $\gamma$ is independant of the variable $x$. Moreover, $\gamma\in \mcd(\Rset)$ since a Gaussian kernel induced an Hilbert space $\msv$ satisfying $|f|_V\leq |f|_\infty+ |\dd f|_\infty  $ for any $f\in \msv$ by \citep[Theorem 9]{glaunes2005transport}.
    %   For the same reason, we have $\phi^v\in \mcd(\Rset^{d+1})$, and thus its last coordinates denoted by $f$ belongs to $\rmC^1(\Rset^{d+1},\Rset^d)$, and by construction $\phi^v=\Psi_\gamma\circ\Phi_f $.
    % \end{proof
    \begin{remark}
      \label{remark:spaceandtime}
      With this choice of kernel, the features associated with the time transformation can be extracted from the momentums $(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0}$ in \eqref{eq:def_v0} by taking the coordinates related to time.
      However, the features related to the space transformation are not only in the space coordinates since the related kernel $K_{\text{space}}$ depends on time as well.
      The kernel's representation has been carefully designed to integrate both space and time, while ensuring that time remains independent of space.
       Initially, we considered separating the spatial and temporal components.
        However, post-hoc analysis of such a representation proved to be challenging.
         The separated spatial and temporal representations are correlated, and understanding this correlation is essential for interpreting the data.
          As a result, concatenating the two representations becomes necessary, though there is no straightforward method for doing so, as they are not commensurable.
           Consequently, we opted for a representation that inherently integrates both space and time.
      \end{remark}
      In \Cref{appendix:kernel_TS_LDDMM}, we give guidelines for selecting the hyperparameters $(\sigma_{T,0},\sigma_{T,1},\sigma_x,c_0,c_1)$.

@@

 Leaving the LDDMM representation, the results of \cite{srivastava2010shape,heo2024logistic} address the representation of curves with the Square-Root Velocity (SRV) representation.
 However, the SRV representation is applied after reparametrization of the temporal dimension of the unit length segment.
 Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
 Very recently, in a functional data analysis (FDA) framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
 However, the space and time representations remain correlated, complicating post-hoc analysis, as discussed in  \Cref{remark:spaceandtime}.
 Additionally, this method is tailored for \textit{continuous objects} and applies only to time series of the \textit{same length}, making the estimation more sensitive to noise. 
 This issue can be addressed through interpolation, but this approach is not always reliable in sparse and irregular sampling scenarios.
 Most FDA approaches, as seen in \cite{shang2014survey,yu2017principal,warmenhoven2021pca}, address this challenge using interpolation or basis function expansion
 In summary, FDA methods typically separate space and time representations for continuous objects, whereas TS-LDDMM algorithm maintain a discrete-to-discrete analysis, inherently integrating both space and time representations.

@@

 All these state-of-the-art methods previously mentionned \cite{glaunes2008large,oh2024stable,wu2024shape,heo2024logistic} are compared to TS-LDDMM in \Cref{appendix: robustness} and \Cref{appendix: shape_classification}.

 Compared to the Metamorphosis framework \cite{blanz2003face}, LDDMM framework has weaker assumptions.
 The 3DMM framework requires that each mesh be re-parametrized into a consistent form where the number of vertices, triangulation, and the anatomical meaning of each vertex are consistent across all meshes, as stated in the introduction of \cite{booth20163d}.
In our context, we do not need such pre-processing; the time series graph can have different sizes.
 \vspace{-1ex}
  \section{Limitations and conclusion}
  \vspace{-1ex}
  \label{sec:limitations}
